Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-10-24 paper/main.tex
remove everything related to blueprint

%###################################################################################################
\section*{Abstract}
\aclp{abm} have become popular tools in the study of complex cellular systems.
However, despite their popularity, it is often difficult to systematically estimate their
parameters and perform advanced analyses, e.g., using the profile-likelihood.
In this work, we study the mechanical properties of rod-shaped bacteria and provide techniques
for estimating their parameters and performing identifiability analysis on them.
Our approach may serve as a blueprint for a modular framework which can be adapted to other
research questions.
We use time series of microscopic images as input data and incorporate advanced effects as the
elasticity of the rod.
The developed model serves as a generalized foundation to describe multiple species of
rod-shaped bacteria such as \textit{E.Coli} or \textit{B.Subtilis}.

@@

In the context of \acp{abm}, however, it is often only shown that the generated results lead to the
desired phenomenon aligning with the observed biological reality.
But this process does not actually estimate the necessary parameters directly \textbf{TODO CITATION}.
Consequently, it reduces the confidence in the constructed models and raises concerns with respect
to the validity of the obtained results.
In this work, we address these challenges using a general-purpose mechanical model for rod-shaped
bacteria as an example.
Our approach provides various methods which can be combined to systematically estimate the
parameters of our model. We deliberately distinguish between conceptually separate components within
this framework.
This modularity highlights a "blueprint" which can be adapted and applied to other \acp{abm}.

@@

\textbf{
    Was mir hier nicht gefällt, ist, dass du von Parameter schätzen bei ABMs kommend ziemlich
    detailliert über Mechanik redest.
    Das passt nicht.
    Man erwartet mehr Infos über den Blueprint und ist verwirrt.
    Worum geht es hier?
}

@@

With this work we have layed out a "blueprint" to estimate the parameters of \acp{abm} using
microscopic image time series.
\textbf{
    Es wird leider überhaupt nicht deutlich, was die Blaupause sein könnte.
    Wenn man das MS liesst, gewinnt man den Eindruck, du machst dies und das, aber ein einheitliches Konzept fehlt.
    Wir müssen zuerst eine Konzeption für das Paper entwickeln, bevor wir über weitere Details reden können.
}

@@

Our work shows that it is possible to estimate the parameters of \acp{abm} on a single-cell level.
Together with the flexible nature of \texttt{cellular\_raza}, this opens up the possibility to
design minimal \acp{abm} which can be fit to data.
As such it formulates a blueprint for the parameter estimation of \acp{abm}.
The developed methods are publicly available as a well-documented python package
(\href{https://github.com/jonaspleyer/cr_mech_coli}{cr\_mech\_coli}).
The individual treatment of the cells allows us in principle to estimate the distribution of the
parameters directly without having to resort to any assumptions on the form of said distribution.
This is true for any parameter, not just the growth rates, since our approach can be generalized to
other individual cellular behaviours.

@@

This work has introduced a novel methodological framework which allows us to estimate the parameters
of models describing rod-shaped bacteria on the single-cell level by using data obtained from
microscopic images.
The approach is agnostic with respect to details of the underlying model and instead only makes use
of the fundamental spatial representation of our bacteria as a collection of vertices.
In this way it highlights a modular blueprint that can be followed by researchers applied to models
of rod-shaped bacteria which incorporate other and more complex behaviours.
Combining these methods, our workflow links experimental data to the parameter estimation of a
mechanistic, computational model and allows us to estimate the individual parameters of the
cell-agents.

Our framework allow researchers to employ well-known methods such as profile-likelihood
\textbf{TODO CITATION} and sensitivity-analysis.
The code is publicly available as the python package
\href{https://github.com/jonaspleyer/cr_mech_coli}{cr\_mech\_coli}.
The utilized cellular properties are contained within the \texttt{cellular\_raza}~\cite{Pleyer2025}
package which provides flexibility in model design, thus enabling researchers to easily reuse,
extend and combine existing building blocks.
This allows researchers to extend the model and incorporate additional biological properties.

>>>>

%###################################################################################################
\section*{Abstract}
\aclp{abm} have become popular tools in the study of complex cellular systems.
However, despite their popularity, it is often difficult to systematically estimate their
parameters and perform advanced analyses, e.g., using the profile-likelihood.
In this work, we study the mechanical properties of rod-shaped bacteria and provide techniques
for estimating their parameters and performing identifiability analysis on them.
Our approach provides a modular framework which can be adapted to a broad range of research
questions regarding models of rod-shaped bacteria.
We use time series of microscopic images as input data and incorporate advanced effects as the
elasticity of the rod.
The developed model serves as a generalized foundation to describe multiple species of
rod-shaped bacteria such as \textit{E.Coli} or \textit{B.Subtilis}.

@@

In the context of \acp{abm}, however, it is often only shown that the generated results lead to the
desired phenomenon aligning with the observed biological reality.
But this process does not actually estimate the necessary parameters directly \textbf{TODO CITATION}.
Consequently, it reduces the confidence in the constructed models and raises concerns with respect
to the validity of the obtained results.
In this work, we address these challenges using a general-purpose mechanical model for rod-shaped
bacteria as an example.
Our approach provides various methods which can be combined to systematically estimate the
parameters of our model. We deliberately distinguish between conceptually separate components within
this framework.

@@

\textbf{
    Was mir hier nicht gefällt, ist, dass du von Parameter schätzen bei ABMs kommend ziemlich
    detailliert über Mechanik redest.
    Das passt nicht.
    Man erwartet mehr Infos über den Blueprint und ist verwirrt.
    Worum geht es hier?
}
\textit{Ich habe den blueprint vollkommen entfernt. Er macht kein Sinn im MS.}

@@

With this work we have layed out how to estimate the parameters of \acp{abm} using
microscopic image time series.
\textbf{
    Es wird leider überhaupt nicht deutlich, was die Blaupause sein könnte.
    Wenn man das MS liesst, gewinnt man den Eindruck, du machst dies und das, aber ein einheitliches Konzept fehlt.
    Wir müssen zuerst eine Konzeption für das Paper entwickeln, bevor wir über weitere Details reden können.
}
\textit{Alles zu blaupause ist wegradiert.}

@@

Our work shows that it is possible to estimate the parameters of \acp{abm} on a single-cell level.
Together with the flexible nature of \texttt{cellular\_raza}, this opens up the possibility to
design minimal \acp{abm} which can be fit to data.
The developed methods are publicly available as a well-documented python package
(\href{https://github.com/jonaspleyer/cr_mech_coli}{cr\_mech\_coli}).
The individual treatment of the cells allows us in principle to estimate the distribution of the
parameters directly without having to resort to any assumptions on the form of said distribution.
This is true for any parameter, not just the growth rates, since our approach can be generalized to
other individual cellular behaviours.

@@

This work has introduced a novel methodological framework which allows us to estimate the parameters
of models describing rod-shaped bacteria on the single-cell level by using data obtained from
microscopic images.
The approach is agnostic with respect to details of the underlying model and instead only makes use
of the fundamental spatial representation of our bacteria as a collection of vertices.
The developed functionalities are modular and can directly be reused by researchers to asses a wide
range of rod-shaped bacterial models.
Combining these methods, our workflow links experimental data to the parameter estimation of a
mechanistic, computational model and allows us to estimate the individual parameters of the
cell-agents.

Our framework allows researchers to employ well-known methods such as profile-likelihood
\textbf{TODO CITATION} and sensitivity-analysis.
The code is publicly available as the python package
\href{https://github.com/jonaspleyer/cr_mech_coli}{cr\_mech\_coli}.
The utilized cellular properties are contained within the \texttt{cellular\_raza}~\cite{Pleyer2025}
package which provides flexibility in model design, thus enabling researchers to easily reuse,
extend and combine existing building blocks.
This allows researchers to extend the model and incorporate additional biological properties.
